HO unification from object language to meta language

### 2024-03-20 paper/appendix.tex
update appendix
appendix meta language

\begin{figure*}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
fcon C ~\Eo~fcon C.
fuva N ~\Eo~fuva N.
flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~

@@

type napp fm -> fm -> o.
napp (fcon C) (fcon C).
napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
napp (fapp[X]) R :- !, napp X R.
napp (fapp A) (fapp B) :- map napp A B.
napp (fuva N) (fuva N).
\end{elpicode}
\caption{Full implementation of the \Eo predicate for \Fo}
\label{code:eo-pred}
\Description[code:eo-pred]{code:eo-pred}
\end{figure*}

\begin{figure*}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
% Congruence
(app A ~\Ue~app B) R S :- fold2 ~(\Ue)~A B R S.
(lam F ~\Ue~lam G) R S :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) R S.
~\Ue~(con C) (con C) S S.
% deref
(uva N A ~\Ue~T) R S :- set? N S F, move F A T1, (T1 ~\Ue~T) R S.
(T ~\Ue~uva N A) R S :- set? N S F, move F A T1, (T ~\Ue~T1) R S.
% flex-flex
(uva N A ~\Ue~uva M B) S S3 :-  unset? M, unset? N,
  distinct A, distinct B, 
  new S W S1, prune W Args1 B Ass, 
  assign N S1 Ass S2, assign M S2 Ass S3.
% assignment
(uva N A ~\Ue~T) R S :- distinct A, not (T = uva _ _), not_occ N S T,  
  bind A T T1, assign N S T1 S1.
(T ~\Ue~uva N A) R S :- distinct A, not (T = uva _ _), not_occ N S T, 
  bind A T T1, assign N S T1 S1.

type distinct list A -> o.
distinct [].
distinct [X|XS] :- name X, not(mem X XS),
distinct XS.

typeabbrev memory A (list (option A)).
type set? nat -> memory A -> A -> o.
set? N S T :- nth N S (some T).
type unset? nat -> memory A -> o.
unset? N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
kind nat type.
type z nat.
type s nat -> nat.
type nth nat -> list A -> A -> o.
nth z [X|_] X.
nth (s N) [_|L] X :- nth N L X.

type new memory A -> nat -> memory A -> o.
new [] z [none].
new [X|XS] (s N) [X|YS] :- new XS N YS.

type prune .
type move .
type beta.
type bind.
type not_occ.
TODO

type fold2 (A -> A1 -> B -> B -> o) -> list A -> list A1 -> B -> B -> o.
fold2 _ [] [] A A.
fold2 F [X|XS] [Y|YS] A A1 :- F X Y A A0, fold2 F XS YS A0 A1.

\end{elpicode}
\caption{Implementation of the \Ue predicate for \Ho}
\label{code:ue-pred}
\Description[code:ue-pred]{code:ue-pred}
\end{figure*}

\begin{figure*}
\begin{elpicode}
type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
fold4 _ [] [] A A B B.
fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.

@@

\end{elpicode}
\caption{Implementation of the compiler}
\label{code:ue-pred}
\Description[code:ue-pred]{code:ue-pred}
\end{figure*}

>>>>

\section{The object language}

\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.                           ~($\customlabel{pred:fequal}{\Eo}$)~
fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
fcon C ~\Eo~fcon C.
fuva N ~\Eo~fuva N.
flam F ~\Eo~T :-                                       ~($\customlabel{clause:eta1}{\eta_l}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => F x ~\Eo~R x.
T ~\Eo~flam F :-                                       ~($\customlabel{clause:eta2}{\eta_r}$)~
  pi x\ beta T [x] (R x), x ~\Eo~x => R x ~\Eo~F x.
fapp [flam X | L] ~\Eo~T :- beta (flam X) L R, R ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_l}$)~
T ~\Eo~fapp [flam X | L] :- beta (flam X) L R, T ~\Eo~R. ~($\customlabel{clause:beta2}{\beta_r}$)~

@@

type napp fm -> fm -> o.
napp (fcon C) (fcon C).
napp (flam F) (flam G) :- pi x\ napp x x => napp (F x) (G x).
napp (fapp[fapp L|M]) R :- !, append L M N, napp (fapp N) R.
napp (fapp[X]) R :- !, napp X R.
napp (fapp A) (fapp B) :- map napp A B.
napp (fuva N) (fuva N).
\end{elpicode}

\section{The meta language}
% \begin{figure*}
\begin{elpicode}
  typeabbrev subst list (option assignment).

  kind inctx type -> type.
  type abs (tm -> inctx A) -> inctx A.
  type val A -> inctx A.

  typeabbrev assignment (inctx tm).

  kind tm  type.
  type app list tm -> tm.
  type lam (tm -> tm) -> tm.
  type con string -> tm.
  type uva  address -> list tm -> tm.

  type (==l) tm -> tm -> subst -> subst ->  o.
  % congruence
  ((app L1) ==l (app L2)) S S1 :- fold2 (==l) L1 L2 S S1.
  ((lam F1) ==l (lam F2)) S S1 :- 
    pi x\ copy x x => ((F1 x) ==l (F2 x)) S S1.
  ((con X) ==l (con X)) S S.
  % set variables
  ((uva N Args) ==l T) S S1 :- 
    mem.set? N S F,!, move F Args T1, (T1 ==l T) S S1.
  (T ==l (uva N Args)) S S1 :- 
    mem.set? N S F,!, move F Args T1, (T ==l T1) S S1.
  % flex-flex
  ((uva M A1) ==l (uva N A2)) S1 S2 :- !, 
    pattern-fragment A1, pattern-fragment A2, 
    prune! M A1 N A2 S1 S2.
  ((uva N Args) ==l T) S S1 :- not_occ N S T, pattern-fragment Args, 
    bind T Args T1, mem.assign N S T1 S1.
  % variable assigment
  (T ==l (uva N Args)) S S1 :- not_occ N S T, pattern-fragment Args, 
    bind T Args T1, mem.assign N S T1 S1.
  (N ==l N) S S :- name N.

  % Note: We suppose the scopes to always be in PF
  type prune! address -> list ho.tm -> address -> 
              list ho.tm -> subst -> subst -> o.
  prune! N A  N A  S S :- !.
  prune! M A  N A  S1 S2 :- !, bind (uva M A) A Ass, 
    mem.assign N S1 Ass S2.
  prune! N A1 N A2 S1 S3 :- !,
    std.assert!(len A1 {len A2}) "Not typechecking", !,
    mem.new S1 W S2, prune-same-variable W A1 A2 [] Ass,
    mem.assign N S2 Ass S3.
  prune! N A1 M A2 S1 S4 :- !,
    mem.new S1 W S2, prune-diff-variables W A1 A2 Ass1 Ass2, 
    mem.assign N S2 Ass1 S3,
    mem.assign M S3 Ass2 S4.

  type prune-same-variable address -> list tm -> list tm -> 
                            list tm -> assignment -> o.
  prune-same-variable N [] [] ACC (val (uva N Args)) :- 
    std.rev ACC Args.
  prune-same-variable N [X|XS] [X|YS] ACC (abs F) :- 
    pi x\ prune-same-variable N XS YS [x|ACC] (F x).
  prune-same-variable N [_|XS] [_|YS] ACC (abs F) :- 
    pi x\ prune-same-variable N XS YS ACC (F x).

  type prune-build-ass1 address -> list tm -> 
                        list bool -> assignment -> o.
  prune-build-ass1 N Acc [] (val (uva N Args)) :- 
    std.rev Acc Args.
  prune-build-ass1 N Acc [tt|L] (abs T) :- 
    pi x\ prune-build-ass1 N [x|Acc] L (T x).
  prune-build-ass1 N Acc [ff|L] (abs T) :- 
    pi x\ prune-build-ass1 N Acc L (T x).

  type build-order list nat -> list tm -> list tm -> o.
  build-order L T R :-
    len L Len, list-init Len z 
      (p\r\ sigma Index Elt\ index L p Index, nth Index T r) R.

  type prune-build-ass2 address -> list tm -> list bool -> 
                      list nat -> assignment -> o.
  prune-build-ass2 N Acc [] Pos (val (uva N Args)) :- 
    std.rev Acc Acc', build-order Pos Acc' Args.
  prune-build-ass2 N Acc [tt|L] Pos (abs T) :- 
    pi x\ prune-build-ass2 N [x|Acc] L Pos (T x).
  prune-build-ass2 N Acc [ff|L] Pos (abs T) :- 
    pi x\ prune-build-ass2 N Acc L Pos (T x).

  type keep list A -> A -> bool -> o.
  keep L A tt :- mem L A, !.
  keep _ _ ff.

  type prune-diff-variables address -> list tm -> list tm -> 
                            assignment -> assignment -> o.
  prune-diff-variables N Args1 Args2 Ass1 Ass2 :-
    std.map Args1 (keep Args2) Bits1, 
    prune-build-ass1 N [] Bits1 Ass1,
    std.map Args2 (keep Args1) Bits2, 
    std.filter Args1 (mem Args2) ToKeep1,
    std.filter Args2 (mem Args1) ToKeep2,
    std.map ToKeep2 (index ToKeep1) Pos,
    prune-build-ass2 N [] Bits2 Pos Ass2.

  type move assignment -> list tm -> tm -> o.
  move (abs Bo)        [H|L] R :- move (Bo H) L R.
  move (val A)         []    A :- !.
  move (val (uva N A)) L     (uva N X) :- std.append A L X.

  type beta tm -> list tm -> tm -> o.
  beta A [] A.
  beta (lam Bo) [H | L] R :- beta (Bo H) L R.
  beta (app A) L (app X) :- std.append A L X.
  beta (uva N A) L (uva N A') :- std.append A L A'.
  beta (con H) L (app [con H | L]).

  type not_occ_aux address -> subst -> tm -> o.
  not_occ_aux N S (uva M _) :- mem.unset? M S, not (N = M).
  not_occ_aux N S (uva M Args) :- mem.set? M S F, 
    move F Args T, not_occ_aux N S T.
  not_occ_aux N S (app L) :- forall1 (not_occ_aux N S) L.
  not_occ_aux N S (lam F) :- pi x\ not_occ_aux N S (F x).
  not_occ_aux _ _ (con _).
  not_occ_aux _ _ X :- name X.

  type not_occ  address -> subst -> tm -> o.
  not_occ N _ (uva N _).
  not_occ N S (uva M Args) :- mem.set? M S F, 
    move F Args T, not_occ N S T.
  not_occ N S (uva M Args) :- mem.unset? M S, 
    std.forall Args (not_occ_aux N S).
  not_occ _ _ (con _).
  not_occ N S (app L) :- not_occ_aux N S (app L).
  not_occ N S (lam L) :- pi x\ not_occ N S (L x).
  not_occ _ _ X :- name X.

  type copy tm -> tm -> o.
  copy (app L)   (app L') :- forall2 copy L L'.
  copy (lam T)   (lam T') :- pi x\ copy x x => copy (T x) (T' x).
  copy (uva N L) (uva N L') :- forall2 copy L L'.
  copy (con C)   (con C).
  copy N N :- name N.

  type bind tm -> list tm -> assignment -> o.
  bind T [] (val T') :- copy T T'.
  bind T [X | TL] (abs T') :- pi x\ copy X x => bind T TL (T' x).

  type deref subst -> tm -> tm -> o.
  deref S X _ :- (var S; var X), halt "flex deref".
  deref H (uva N L)  X         :- mem.set? N H T, 
    move T L X', !, deref H X' X.
  deref H (app L)   (app L1)   :- forall2 (deref H) L L1.
  deref _ (con X)   (con X).
  deref H (uva X L) (uva X L1) :- mem.unset? X H, 
    forall2 (deref H) L L1.
  deref H (lam F)   (lam G)    :- pi x\ deref H (F x) (G x).
  deref _ N         N          :- name N.

  type deref-assmt subst -> assignment -> assignment -> o.
  deref-assmt S (abs T) (abs R) :- pi x\ deref-assmt S (T x) (R x).
  deref-assmt S (val T) (val R) :- deref S T R.
\end{elpicode}
% \caption{Implementation of the \Ue predicate for \Ho}
% \label{code:ue-pred}
% \Description[code:ue-pred]{code:ue-pred}
% \end{figure*}

\section{The compiler}
% \begin{figure*}
% \caption{Implementation of the compiler}
% \label{code:ue-pred}
% \Description[code:ue-pred]{code:ue-pred}
% \end{figure*}

\section{The progress function}

\section{The decompiler}

\section{Auxiliary functions}

\begin{elpicode}
  type fold4 (A -> A1 -> B -> B -> C -> C -> o) -> list A -> list A1 -> B -> B -> C -> C -> o.
  fold4 _ [] [] A A B B.
  fold4 F [X|XS] [Y|YS] A A1 B B1 :- F X Y A A0 B B0, fold4 F XS YS A0 A1 B0 B1.

@@

  \end{elpicode}
